---
title: "Plotting Exp and Obs Geno Freqs"
output: 
  html_notebook:
  toc: true
  toc_float: true
---



We are going to go through the different data sets here and just make
fairly standardized figures of what the expected and observed genotype
frequencies look like.

Load up some libraries.
```{r}
library(tidyverse)
library(stringr)
library(genoscapeRtools)
```

Get the meta data:
```{r}
meta_data <- read_rds("processed_data/meta_data.rds")
```

## A function to grab the data and make the plots

This will be a straightforward thing where we give it the "name" of the data set
and it reads it in and then does the calcs and saves the plot as a pdf in the 
outputs directory (if pdf == TRUE).  It will scale the height
appropriately if there are a lot of populations.

```{r}
# first source a function that makes it easy to compute the boundaries
# of the geno freqs
source("R/geno-freq-boundaries.R")
plot_geno_freqs <- function(name, alpha = 0.2, plot_all_pops = FALSE) {
  message("starting ", name)
  d012 <- genoscapeRtools::read_012(prefix = file.path("processed_data", name), gz = TRUE)
  pops <- readr::read_csv(paste0("processed_data/", name, "_pops.csv"))
  
  # this grabs the biggest sample group
  big_pop <- pops %>%
    count(pop) %>%
    top_n(n = 1, wt = n) %>%
    .$pop %>%
    .[1]
  
  # this gets the sample names in that biggest pop
  bp_pops <- pops %>%
    filter(pop == big_pop) 
  
  bp_names <- bp_pops %>%
    .$sample
  
  # this just takes data from the biggest pop samples
  bp012 <- d012[bp_names,]
  
  gfc <- genoscapeRtools::geno_freq_calcs(bp012, bp_pops) %>%
    filter(!is.nan(z_score))  # this chucks out monomorphic loci
  
  
  gbp <- ggplot(gfc , aes(x = p_exp, y = p_obs, colour = geno)) +
    geom_jitter(alpha = alpha, position = position_jitter(width = 0.01, height = 0.01)) +
    facet_grid(pop ~ geno) +
    geom_polygon(data = geno_freq_boundaries(), fill = NA, linetype = "dashed", colour = "black") + 
    geom_abline(slope = 1, intercept = 0, linetype = "solid") +
    ggtitle(paste0(name, " (", bp_pops$pop[1],"), ",  "n = ", nrow(bp_pops)))
  
  if (!file.exists("outputs")) dir.create("outputs")
  ggsave(gbp, filename = paste0("outputs/", name, "_big_pop.pdf"), width = 8, height = 3)
  
  big_pop  # return the name of the population with the largest sample
}
```

## Running it over everyone

This can be done easily
```{r}
big_pops <- lapply(1:nrow(meta_data), function(i) plot_geno_freqs(meta_data$names[i]))
```

And once we have done that, we can record which populations were the "big pops" (the largest samples)
for all the different groups:
```{r}
meta_with_big_pops <- meta_data %>%
  mutate(big_pop = unlist(big_pops)) %>%
  select(study, names, big_pop, everything())

write_rds(meta_with_big_pops, path = "processed_data/meta_with_big_pops.rds")
```

Now we just have to estimate the genotyping error rates from each of these....